National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-19 Manuscript.Rmd
revised manuscript, bib

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

@@

Table 1 about here:
\[
\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Māori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Pol Orient &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule
\end{tabular}
}
\end{table}
\]

## Demographic Indicators

@@

### Ethnicity
Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Māori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

@@

## Equation 1
\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Māori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
\end{aligned}
\]

@@

### Missing data
Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Details of this analysis can be found at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("Figs/USE_2_coefficientplots-1.png")
```

@@

\centering 
\caption{"Longitudinal Models"}\label{}

@@

 & Personal Well-Being & Life Satisfaction \\

@@

Employed1                          & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\

@@

EthnicCatsMāori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\

@@

Partner1                           & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\

@@

@@

  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
  - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of identifying as Māori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Māori group tend to have lower Life Satisfaction.
  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("Figs/USElifsat_expected-1.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799). Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

@@

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("Figs/predicted_PWI_impute-1.png")
```

@@

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("Figs/predicted_lifesat_imputed-1.png")
```

@@

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y}$ = 6.69, se = 0.02,$\Delta \hat{y} = .21$). To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner. By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$). In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner. By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).

@@

Two fundamental question arise from the patterns we observe in this study. The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics? During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [CITE]. Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [CITE, CITE]. Previous research finds an increase in traditional religious affiliation in the wake of the Christchurch earthquakes, with no immediate effect in subjective well-being. Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [CITE]. Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster. On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean. Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations. Despite the longitudinal quality of this study, a causal explanation remains elusive.

>>>>

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","graphicx"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

@@

Table 1 about here:
\[
\begin{table}[ ht ] 
\centering 
\caption{}\label{}
\scalebox{.6}{
\begin{tabular}{ l c c c c c c c c c }
\toprule
 &   \multicolumn{ 9 }{c}{ Wave }\\ 
  & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018 \\ 
 & n = \emph{missing} & n = \emph{missing} & n = 8547 & n = 13472 & n = \emph{missing} & n = 11687 & n = 16048 & n = 14174 & n = 14883 \\ 
 \midrule
Age &   &   &   &   &   &   &   &   &  \\ 
NZdep &   &   &   &   &   &   &   &   &  \\ 
Education &   &   &   &   &   &   &   &   &  \\ 
Employed &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2094 (24.5\%) & 2955 (21.9\%) &  (\%) & 2525 (21.6\%) & 3486 (21.7\%) & 3223 (22.7\%) & 3404 (22.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6453 (75.5\%) & 10517 (78.1\%) &  (\%) & 9162 (78.4\%) & 12562 (78.3\%) & 10951 (77.3\%) & 11479 (77.1\%)\\ 
EthnicCategories &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Euro &  (\%) &  (\%) & 6731 (78.8\%) & 11024 (81.8\%) &  (\%) & 9671 (82.8\%) & 13389 (83.4\%) & 11814 (83.3\%) & 12427 (83.5\%)\\ 
\hspace{6pt}    Maori &  (\%) &  (\%) & 1188 (13.9\%) & 1608 (11.9\%) &  (\%) & 1341 (11.5\%) & 1749 (10.9\%) & 1620 (11.4\%) & 1677 (11.3\%)\\ 
\hspace{6pt}    Pacific &  (\%) &  (\%) & 284 (3.3\%) & 339 (2.5\%) &  (\%) & 277 (2.4\%) & 331 (2.1\%) & 243 (1.7\%) & 261 (1.8\%)\\ 
\hspace{6pt}    Asian &  (\%) &  (\%) & 344 (4\%) & 501 (3.7\%) &  (\%) & 398 (3.4\%) & 579 (3.6\%) & 497 (3.5\%) & 518 (3.5\%)\\ 
Male &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 5302 (62\%) & 8457 (62.8\%) &  (\%) & 7325 (62.7\%) & 10148 (63.2\%) & 9011 (63.6\%) & 9433 (63.4\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 3245 (38\%) & 5015 (37.2\%) &  (\%) & 4362 (37.3\%) & 5900 (36.8\%) & 5163 (36.4\%) & 5450 (36.6\%)\\ 
Partner &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    0 &  (\%) &  (\%) & 2408 (28.2\%) & 3546 (26.3\%) &  (\%) & 2815 (24.1\%) & 3773 (23.5\%) & 3330 (23.5\%) & 3556 (23.9\%)\\ 
\hspace{6pt}    1 &  (\%) &  (\%) & 6139 (71.8\%) & 9926 (73.7\%) &  (\%) & 8872 (75.9\%) & 12275 (76.5\%) & 10844 (76.5\%) & 11327 (76.1\%)\\ 
Pol Orient &   &   &   &   &   &   &   &   &  \\ 
Urban &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    Not\_Urban &  (\%) &  (\%) & 2865 (33.5\%) & 4278 (31.8\%) &  (\%) & 3951 (33.8\%) & 5538 (34.5\%) & 2610 (18.4\%) & 2661 (17.9\%)\\ 
\hspace{6pt}    Urban &  (\%) &  (\%) & 5682 (66.5\%) & 9194 (68.2\%) &  (\%) & 7736 (66.2\%) & 10510 (65.5\%) & 11564 (81.6\%) & 12222 (82.1\%)\\ 
Beliefs &   &   &   &   &   &   &   &   &  \\ 
\hspace{6pt}    \_Skeptic\_ &  (\%) &  (\%) & 2048 (24\%) & 3286 (24.4\%) &  (\%) & 2951 (25.3\%) & 4116 (25.6\%) & 3916 (27.6\%) & 4120 (27.7\%)\\ 
\hspace{6pt}    \_SpiritExcludesGod\_ &  (\%) &  (\%) & 2295 (26.9\%) & 3735 (27.7\%) &  (\%) & 3169 (27.1\%) & 4501 (28\%) & 3918 (27.6\%) & 3970 (26.7\%)\\ 
\hspace{6pt}    GodAndSpirit &  (\%) &  (\%) & 3707 (43.4\%) & 5499 (40.8\%) &  (\%) & 5119 (43.8\%) & 6453 (40.2\%) & 5464 (38.5\%) & 6210 (41.7\%)\\ 
\hspace{6pt}    GodExcludesSpirit &  (\%) &  (\%) & 497 (5.8\%) & 952 (7.1\%) &  (\%) & 448 (3.8\%) & 978 (6.1\%) & 876 (6.2\%) & 583 (3.9\%)\\ 
LIFESAT &   &   &   &   &   &   &   &   &  \\ 
PWI &   &   &   &   &   &   &   &   &  \\ 
\bottomrule
\end{tabular}
}
\end{table}
\]

## Demographic Indicators
We included standard demographic indicators used in New Zealand Attitudes and Values Research: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf)

@@

### Ethnicity
Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

@@

## Equation 1
\[
\begin{aligned}
\operatorname{{\small LIFESAT/PWI}}  &\sim N \left(\mu,\sigma^2 \right) \\ \mu &=\alpha_{j[i]} + \beta_{1}(\operatorname{Years}) + \beta_{2}(\operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}})\ + \\
&\quad \beta_{3}(\operatorname{Beliefs}_{\operatorname{GodAndSpirit}}) + \beta_{4}(\operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) + \beta_{5}(\operatorname{Age.10yrs.C})\ + \\
&\quad \beta_{6}(\operatorname{Deprivation.S}) + \beta_{7}(\operatorname{Edu.S}) + \beta_{8}(\operatorname{Employed})\ + \\
&\quad \beta_{9}(\operatorname{EthnicCats}_{\operatorname{Maori}}) + \beta_{10}(\operatorname{EthnicCats}_{\operatorname{Pacific}}) + \beta_{11}(\operatorname{EthnicCats}_{\operatorname{Asian}})\ + \\
&\quad \beta_{12}(\operatorname{Male}_{\operatorname{1}}) + \beta_{13}(\operatorname{Partner}_{\operatorname{Has\ }}) + \beta_{14}(\operatorname{Pol.Orient.S})\ + \\
&\quad \beta_{15}(\operatorname{Urban}) + \beta_{16}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{SpiritExcludesGod}}) + \beta_{17}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodAndSpirit}})\ + \\
&\quad \beta_{18}(\operatorname{Years} \times \operatorname{Beliefs}_{\operatorname{GodExcludesSpirit}}) \\ \alpha_{j} &\sim N \left(\mu_{\alpha_{j}},\sigma^2_{\alpha_{j}} \right) , \operatorname{ for  Id }~j~= 1, \dots~J = 20,979~\operatorname{Individuals}
\end{aligned}
\]

@@

### Missing data
Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N >3 to N = 9 waves. Additionally, the findings reported below were robust to modeling only the effects of years and beliefs, without adjusting for the standard NZAVS demographic covariates. The supplement describing these analysis is located at: https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing.

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")
```

@@

\begin{center}

@@

 & PWI & Life Satisfaction \\

@@

EmployedEmployed                   & $\mathbf{0.09}^{***}$  & $\mathbf{0.08}^{***}$  \\

@@

EthnicCatsMaori                    & $\mathbf{-0.30}^{***}$ & $\mathbf{-0.06}^{**}$  \\

@@

PartnerHas Partner                 & $\mathbf{0.61}^{***}$  & $\mathbf{0.42}^{***}$  \\

@@

\caption{Regression Models}
\label{table:coefficients}
\end{center}

@@

  - The effect of years is not reliable (beta = -5.50e-04, SE = 2.69e-03, std. beta = -9.32e-04, p = 0.838).  We infer that life satisfaction did not increase in the Skeptic population between 2010 and 2018.
  - The effect of beliefs in *Spirit Excludes God* is reliably negative (beta = -0.08, SE = 0.02, std. beta = 0.01, p < .001). We infer that overall the population that believes in a Spirit but does not believe in a God is expected to have a lower Life Satisfaction than skeptics. 
  - The effect of beliefs in *God And Spirit* is not reliable (beta = -0.04, SE = 0.02, std. beta = 0.04, p = 0.105). People who believe in God but doubt a Spirit/Life force are expected to have had similar levels of life satisfaction in Wave 10 as skeptics.
  - The effect of beliefs in *God Excludes Spirit* is reliably positive (beta = 0.07, SE = 0.04, std. beta = 0.05, p = 0.062).We infer that overall the population that believes in a God but does not believe in a spirit is not reliably different from  the Skeptics population in Life Satisfaction. 
  - The effect of Age.10yrs.C is reliably positive (beta = 0.09, SE = 5.26e-03, std. beta = 0.10, p < .001). Older people have greater Life Satisfaction
  - The effect of Deprivation (S) is  is reliably negative (beta = -0.07, SE = 5.50e-03, std. beta = -0.06, p < .001). People who are more deprived have lower Life Satisfaction.
  - The effect of Education (S) is reliably positive (beta = 0.06, SE = 6.33e-03, std. beta = 0.05, p < .001). Educated people have greater Life Satisfaction
  - The effect of being Employed is reliably positive (beta = 0.08, SE = 9.69e-03, std. beta = 0.07, p < .001). People who are employed tend to have greater Life Satisfaction
  - The effect of identifying as Maori is reliably negative (beta = -0.06, SE = 0.02, std. beta = -0.05, p < .01).  People who belong to the minority Maori group tend to have lower Life Satisfaction.
  - The effect of identifying as Pacific is reliably negative (beta = -0.08, SE = 0.04, std. beta = -0.07, p < .05).  People who belong to the minority Pacific peoples group tend to have lower life satisfaction.
  - The effect of identifying as Asian is not reliably different from people of European ethnicity in life satisfaction(beta = -0.04, SE = 0.03, std. beta = -0.03, p = 0.223).
  - The effect of identifying as male is reliably negative (beta = -0.19, SE = 0.01, std. beta = -0.16, p < .001).
  - The effect of having a partner is reliably positive (beta = 0.42, SE = 0.01, std. beta = 0.35, p < .001).
  - The effect of political conservativism (S) is reliably positive (beta = 0.01, SE = 4.23e-03, std. beta = 0.01, p < .01).
  - The effect of living in an urban area is reliably negative (beta = -0.02, SE = 8.85e-03, std. beta = -0.02, p < .05).

@@

```{r fig.cap = "Prediction Plot Life Satisfaction Spirit Belief Longitudinal Models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USElifsat_expected-1.png")
```
  - The interaction of years $\times$ beliefs in *Spirit Excludes God* ludesGod is reliably positive (beta = 0.02, SE = 3.66e-03, std. beta = 0.03, p < .001). We infer that, unlike the Skeptic population, people with beliefs in a Spirit or Life Force who do not believe in a God tend to grow in their Life Satisfaction over time.
  - The interaction of years $\times$ *God And Spirit*  is reliably positive (beta = 0.01, SE = 3.32e-03, std. beta = 0.02, p < .001). We infer that people who believe in a spirit or life force who also believe in a God tend to grow in their life satisfaction over time
  - The effect of years $\times$ beliefs in *God Excludes Spirit*  is not reliable (beta = -1.59e-03, SE = 6.25e-03, std. beta = -2.70e-03, p = 0.799). Thus, God believers who doubt in a Spirit or Life Force start maintain higher Life-Satisfaction than do other groups, but they are not expected to grow in Life-Satisfaction over time.

@@

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_PWI_impute-1.png")
```

@@

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/predicted_lifesat_imputed-1.png")
```

@@

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$). To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner. By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$). In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner. By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).

@@

Two fundamental question arise from the patterns we observe in this study. The first question is why those who believed in Spirit or Life Force were initially lower in their personal well-being and life-satisfaction in the initial years of this study in comparison to religious and non-religious skeptics? During the early years of New Zealand Attitudes and Values data collection New Zealand was emerging from the twin impacts of a global financial recession and a series of devastating earthquakes that destroyed 30% of Christchurch, New Zealand's second largest city [@greaves2015regional]. Orthogonal research observes that economic and ecological distress are associated with increases in traditional religious beliefs [@sibley2012faith]. Experimental research has found among religious people death anxiety provokes acceptance of non-traditional religious beliefs [@norenzayan2006belief]. Because spiritual belief items were only introduced during the year of the earthquakes, we cannot here resolve whether the shifting relationship between spirit beliefs and well-being were the result of demand for spiritual beliefs among those secular and religious people who were adversely affected by the financial crisis or natural disaster. On the other hand, such exogenous influences might have adversely affected those with Spirit/Life force beliefs in the early phases of this study, and the patterns observed here are better understood as a regression to a population mean. Though beyond the scope of this study, how subjective well-being changes in response to changes in spiritual beliefs, as well as special sensitivities to distress among people with spiritual beliefs are important questions for future investigations. Despite the longitudinal quality of this study, a causal explanation remains elusive.
